---
title: "ATAC-seq normalization summary, polished, permuted GC content, normalized scores"
author: "Koen Van den Berge"
date: "12/12/2019"
output: html_document
---

This script only removes a few methods from the benchmark that are used for other purposes, e.g. robustness benchmarks. 

```{r}
library(tidyverse)
## convert scores to [0,1] scale: respects difference magnitude better as compared to ranks
getMeanScaledScore <- function(scores){
  # the more positive, the better
  scoreT <- t(scores)
  ranked_scores <- apply(scoreT,1, function(x){
    x2 <- x - min(x)
    x3 <- x2 / max(x2)
    return(x3) #in [0,1]
  })
  mean_score_rank <- rowMeans(ranked_scores)
  mean_score_rank <- mean_score_rank[order(mean_score_rank, decreasing=TRUE)]
  return(mean_score_rank)
}

rmMethods <- c("gcqn_mean", "gcqn_median_10", "gcqn_median_20",  "gcqn_median_100",
               "gcqn_median_permuted")

library(ggplot2)
files <- list.files("~/Dropbox/research/atacseq/bulk/analysis/benchmarkNormalization/permuted/",
                    pattern = "*scores_permuted.rds$")
datasets <- unlist(lapply(strsplit(files, split="_"),"[[",1))
scores <- sapply(1:length(files), function(ii) readRDS(files[ii]))
# remove other methods
scores <- lapply(scores, function(scoreMat){
  scoreNames <- rownames(scoreMat)
  rmID <- unlist(sapply(rmMethods, function(method){
    grep(x=scoreNames, pattern=method)
  }))
  return(scoreMat[-rmID,])
})

pal <- wesanderson::wes_palette("Zissou1", n=40, type="continuous")
```

# Conditional on the dimension of unwanted variation latent variables: what performs best?

Here we stratify performances based on the number of estimated unwanted variation variables that was adjusted for after the normalization.


## No latent variables

```{r}
scoresNoUV <- lapply(scores, function(scoresDataset){
  scoresDataset[grep(x=rownames(scoresDataset), pattern="no_uv"),]
})
scoresNoUVScaled <- lapply(scoresNoUV, getMeanScaledScore)
scoreVectorNoUV <- do.call(c,scoresNoUVScaled)
scoreDf0 <- data.frame(method=names(scoreVectorNoUV),
                          dataset=unlist(mapply(rep,datasets,unlist(lapply(scoresNoUVScaled,length)))),
                          rank=scoreVectorNoUV)
scoreDf0$methodShort <- unlist(lapply(strsplit(as.character(scoreDf0$method), split=","), "[[", 2))
scoreDf0$methodShort[scoreDf0$methodShort == "gcqn_median_50"] <- "GC-QN"
scoreDf0$methodShort[scoreDf0$methodShort == "edaseq"] <- "EDASeq"
scoreDf0$methodShort[scoreDf0$methodShort == "gcqn_smooth"] <- "GC-QN_smooth"
scoreDf0$methodShort[scoreDf0$methodShort == "fq"] <- "FQ"
scoreDf0$methodShort[scoreDf0$methodShort == "tmm"] <- "TMM"
scoreDf0$methodShort[scoreDf0$methodShort == "deseq"] <- "DESeq2"
scoreDf0$methodShort[scoreDf0$methodShort == "uq"] <- "UQ"
scoreDf0$methodShort[scoreDf0$methodShort == "none"] <- "None"
scoreDf0$methodShort[scoreDf0$methodShort == "sum"] <- "Sum"
scoreDf0$methodShort <- factor(scoreDf0$methodShort)

## scale ranks
scoreDf0 <- scoreDf0 %>% group_by(dataset) #%>% mutate(scaledRank=scale(rank))

## take average of scaled ranks
avRank0 <- scoreDf0 %>% group_by(methodShort) %>% summarize(avRank=mean(rank))
oo0 <- order(avRank0$avRank, decreasing=TRUE)
orderedMethodShort0 <- avRank0$methodShort[oo0]

scoreDfOrdered0 <- scoreDf0
scoreDfOrdered0$methodShort <- factor(scoreDfOrdered0$methodShort, levels=orderedMethodShort0)
ggplot(scoreDfOrdered0, aes(y=methodShort, x=dataset, fill=rank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Normalized Scone\nscore') +
  ggtitle("Without RUV variables") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_noRUV_permuted.pdf")
```

## One latent variable

```{r}
scoresUV1 <- lapply(scores, function(scoresDataset){
  scoresDataset[grep(x=rownames(scoresDataset), pattern="ruv_k=1"),]
})
scoresUV1Scaled <- lapply(scoresUV1, getMeanScaledScore)
scoreVectorUV1 <- do.call(c,scoresUV1Scaled)
scoreDf1 <- data.frame(method=names(scoreVectorUV1),
                          dataset=unlist(mapply(rep,datasets,unlist(lapply(scoresUV1Scaled,length)))),
                          rank=scoreVectorUV1)
scoreDf1$methodShort <- unlist(lapply(strsplit(as.character(scoreDf1$method), split=","), "[[", 2))
scoreDf1$methodShort[scoreDf1$methodShort == "gcqn_median_50"] <- "GC-QN"
scoreDf1$methodShort[scoreDf1$methodShort == "edaseq"] <- "EDASeq"
scoreDf1$methodShort[scoreDf1$methodShort == "gcqn_smooth"] <- "GC-QN_smooth"
scoreDf1$methodShort[scoreDf1$methodShort == "fq"] <- "FQ"
scoreDf1$methodShort[scoreDf1$methodShort == "tmm"] <- "TMM"
scoreDf1$methodShort[scoreDf1$methodShort == "deseq"] <- "DESeq2"
scoreDf1$methodShort[scoreDf1$methodShort == "uq"] <- "UQ"
scoreDf1$methodShort[scoreDf1$methodShort == "none"] <- "None"
scoreDf1$methodShort[scoreDf1$methodShort == "sum"] <- "Sum"
scoreDf1$methodShort <- factor(scoreDf1$methodShort)

## scale ranks
scoreDf1 <- scoreDf1 %>% group_by(dataset) 

avRank1 <- scoreDf1 %>% group_by(methodShort) %>% summarize(avRank=mean(rank))
oo1 <- order(avRank1$avRank, decreasing=TRUE)
orderedMethodShort1 <- avRank1$methodShort[oo1]

scoreDfOrdered1 <- scoreDf1
scoreDfOrdered1$methodShort <- factor(scoreDfOrdered1$methodShort, levels=orderedMethodShort1)
ggplot(scoreDfOrdered1, aes(y=methodShort, x=dataset, fill=rank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Normalized Scone\nscore') +
  ggtitle("One RUV variable") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_RUV1_permuted.pdf")
```

## Two latent variables

```{r}
scoresUV2 <- lapply(scores, function(scoresDataset){
  scoresDataset[grep(x=rownames(scoresDataset), pattern="ruv_k=2"),]
})
scoresUV2Scaled <- lapply(scoresUV2, getMeanScaledScore)
scoreVectorUV2 <- do.call(c,scoresUV2Scaled)
scoreDf2 <- data.frame(method=names(scoreVectorUV2),
                          dataset=unlist(mapply(rep,datasets,unlist(lapply(scoresUV2Scaled,length)))),
                          rank=scoreVectorUV2)
scoreDf2$methodShort <- unlist(lapply(strsplit(as.character(scoreDf2$method), split=","), "[[", 2))
scoreDf2$methodShort[scoreDf2$methodShort == "gcqn_median_50"] <- "GC-QN"
scoreDf2$methodShort[scoreDf2$methodShort == "edaseq"] <- "EDASeq"
scoreDf2$methodShort[scoreDf2$methodShort == "gcqn_smooth"] <- "GC-QN_smooth"
scoreDf2$methodShort[scoreDf2$methodShort == "fq"] <- "FQ"
scoreDf2$methodShort[scoreDf2$methodShort == "tmm"] <- "TMM"
scoreDf2$methodShort[scoreDf2$methodShort == "deseq"] <- "DESeq2"
scoreDf2$methodShort[scoreDf2$methodShort == "uq"] <- "UQ"
scoreDf2$methodShort[scoreDf2$methodShort == "none"] <- "None"
scoreDf2$methodShort[scoreDf2$methodShort == "sum"] <- "Sum"
scoreDf2$methodShort <- factor(scoreDf2$methodShort)

## scale ranks
scoreDf2 <- scoreDf2 %>% group_by(dataset) 

avRank2 <- scoreDf2 %>% group_by(methodShort) %>% summarize(avRank=mean(rank))
oo2 <- order(avRank2$avRank, decreasing=TRUE)
orderedMethodShort2 <- avRank2$methodShort[oo2]

scoreDfOrdered2 <- scoreDf2
scoreDfOrdered2$methodShort <- factor(scoreDfOrdered2$methodShort, levels=orderedMethodShort2)
ggplot(scoreDfOrdered2, aes(y=methodShort, x=dataset, fill=rank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Normalized Scone\nscore') +
  ggtitle("Two RUV variables") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_RUV2_permuted.pdf")
```

## Three latent variables

```{r}
scoresUV3 <- lapply(scores, function(scoresDataset){
  scoresDataset[grep(x=rownames(scoresDataset), pattern="ruv_k=3"),]
})
scoresUV3Scaled <- lapply(scoresUV3, getMeanScaledScore)
scoreVectorUV3 <- do.call(c,scoresUV3Scaled)
scoreDf3 <- data.frame(method=names(scoreVectorUV3),
                          dataset=unlist(mapply(rep,datasets,unlist(lapply(scoresUV3Scaled,length)))),
                          rank=scoreVectorUV3)
scoreDf3$methodShort <- unlist(lapply(strsplit(as.character(scoreDf3$method), split=","), "[[", 2))
scoreDf3$methodShort[scoreDf3$methodShort == "gcqn_median_50"] <- "GC-QN"
scoreDf3$methodShort[scoreDf3$methodShort == "edaseq"] <- "EDASeq"
scoreDf3$methodShort[scoreDf3$methodShort == "gcqn_smooth"] <- "GC-QN_smooth"
scoreDf3$methodShort[scoreDf3$methodShort == "fq"] <- "FQ"
scoreDf3$methodShort[scoreDf3$methodShort == "tmm"] <- "TMM"
scoreDf3$methodShort[scoreDf3$methodShort == "deseq"] <- "DESeq2"
scoreDf3$methodShort[scoreDf3$methodShort == "uq"] <- "UQ"
scoreDf3$methodShort[scoreDf3$methodShort == "none"] <- "None"
scoreDf3$methodShort[scoreDf3$methodShort == "sum"] <- "Sum"
scoreDf3$methodShort <- factor(scoreDf3$methodShort)

## scale ranks
scoreDf3 <- scoreDf3 %>% group_by(dataset)

avRank3 <- scoreDf3 %>% group_by(methodShort) %>% summarize(avRank=mean(rank))
oo3 <- order(avRank3$avRank, decreasing=TRUE)
orderedMethodShort3 <- avRank3$methodShort[oo3]

scoreDfOrdered3 <- scoreDf3
scoreDfOrdered3$methodShort <- factor(scoreDfOrdered3$methodShort, levels=orderedMethodShort3)
ggplot(scoreDfOrdered3, aes(y=methodShort, x=dataset, fill=rank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Normalized Scone\nscore') +
  ggtitle("Three RUV variables") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_RUV3_permuted.pdf")

```

## Four latent variables

```{r}
scoresUV4 <- lapply(scores, function(scoresDataset){
  scoresDataset[grep(x=rownames(scoresDataset), pattern="ruv_k=4"),]
})
scoresUV4Scaled <- lapply(scoresUV4, getMeanScaledScore)
scoreVectorUV4 <- do.call(c,scoresUV4Scaled)
scoreDf4 <- data.frame(method=names(scoreVectorUV4),
                          dataset=unlist(mapply(rep,datasets,unlist(lapply(scoresUV4Scaled,length)))),
                          rank=scoreVectorUV4)
scoreDf4$methodShort <- unlist(lapply(strsplit(as.character(scoreDf4$method), split=","), "[[", 2))
scoreDf4$methodShort[scoreDf4$methodShort == "gcqn_median_50"] <- "GC-QN"
scoreDf4$methodShort[scoreDf4$methodShort == "edaseq"] <- "EDASeq"
scoreDf4$methodShort[scoreDf4$methodShort == "gcqn_smooth"] <- "GC-QN_smooth"
scoreDf4$methodShort[scoreDf4$methodShort == "fq"] <- "FQ"
scoreDf4$methodShort[scoreDf4$methodShort == "tmm"] <- "TMM"
scoreDf4$methodShort[scoreDf4$methodShort == "deseq"] <- "DESeq2"
scoreDf4$methodShort[scoreDf4$methodShort == "uq"] <- "UQ"
scoreDf4$methodShort[scoreDf4$methodShort == "none"] <- "None"
scoreDf4$methodShort[scoreDf4$methodShort == "sum"] <- "Sum"
scoreDf4$methodShort <- factor(scoreDf4$methodShort)

## scale ranks
scoreDf4 <- scoreDf4 %>% group_by(dataset)

avRank4 <- scoreDf4 %>% group_by(methodShort) %>% summarize(avRank=mean(rank))
oo4 <- order(avRank4$avRank, decreasing=TRUE)
orderedMethodShort4 <- avRank4$methodShort[oo4]

scoreDfOrdered4 <- scoreDf4
scoreDfOrdered4$methodShort <- factor(scoreDfOrdered4$methodShort, levels=orderedMethodShort4)
ggplot(scoreDfOrdered4, aes(y=methodShort, x=dataset, fill=rank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Normalized Scone\nscore') +
  ggtitle("Four RUV variables") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_RUV4_permuted.pdf")

```

## Five latent variables

```{r}
scoresUV5 <- lapply(scores, function(scoresDataset){
  scoresDataset[grep(x=rownames(scoresDataset), pattern="ruv_k=5"),]
})
scoresUV5Scaled <- lapply(scoresUV5, getMeanScaledScore)
scoreVectorUV5 <- do.call(c,scoresUV5Scaled)
scoreDf5 <- data.frame(method=names(scoreVectorUV5),
                          dataset=unlist(mapply(rep,datasets,unlist(lapply(scoresUV5Scaled,length)))),
                          rank=scoreVectorUV5)
scoreDf5$methodShort <- unlist(lapply(strsplit(as.character(scoreDf5$method), split=","), "[[", 2))
scoreDf5$methodShort[scoreDf5$methodShort == "gcqn_median_50"] <- "GC-QN"
scoreDf5$methodShort[scoreDf5$methodShort == "edaseq"] <- "EDASeq"
scoreDf5$methodShort[scoreDf5$methodShort == "gcqn_smooth"] <- "GC-QN_smooth"
scoreDf5$methodShort[scoreDf5$methodShort == "fq"] <- "FQ"
scoreDf5$methodShort[scoreDf5$methodShort == "tmm"] <- "TMM"
scoreDf5$methodShort[scoreDf5$methodShort == "deseq"] <- "DESeq2"
scoreDf5$methodShort[scoreDf5$methodShort == "uq"] <- "UQ"
scoreDf5$methodShort[scoreDf5$methodShort == "none"] <- "None"
scoreDf5$methodShort[scoreDf5$methodShort == "sum"] <- "Sum"
scoreDf5$methodShort <- factor(scoreDf5$methodShort)

## scale ranks
scoreDf5 <- scoreDf5 %>% group_by(dataset)

avRank5 <- scoreDf5 %>% group_by(methodShort) %>% summarize(avRank=mean(rank))
oo5 <- order(avRank5$avRank, decreasing=TRUE)
orderedMethodShort5 <- avRank5$methodShort[oo5]

scoreDfOrdered5 <- scoreDf5
scoreDfOrdered5$methodShort <- factor(scoreDfOrdered5$methodShort, levels=orderedMethodShort5)
ggplot(scoreDfOrdered5, aes(y=methodShort, x=dataset, fill=rank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Normalized Scone\nscore') +
  ggtitle("Five RUV variables") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_RUV5_permuted.pdf")
```




