---
title: "Liu 2019 ATAC-seq atlas case study"
author: "Koen Van den Berge"
date: "5/19/2020"
output: html_document
---

```{r}
gcqn_qsmooth <- function(counts, gcGroups, bio){
  gcBinNormCounts <- matrix(NA, nrow=nrow(counts), ncol=ncol(counts), dimnames=list(rownames(counts),colnames(counts)))
  for(ii in 1:nlevels(gcGroups)){
    id <- which(gcGroups==levels(gcGroups)[ii])
    countBin <- counts[id,]
    qs <- qsmooth(countBin, group_factor=bio)
    normCountBin <- qs@qsmoothData
    normCountBin <- round(normCountBin)
    normCountBin[normCountBin<0] <- 0
    gcBinNormCounts[id,] <- normCountBin
  }
  return(gcBinNormCounts)
}
library(GenomicAlignments)
library(rafalib)
library(cqn)
library(Biobase)
library(edgeR)
library(mclust)
library(umap)
source("../../../methods/gcqn_validated.R")
library(tidyverse)
library(qsmooth)
library(RUVSeq)
library(RColorBrewer)
plotGCHex <- function(gr, counts){
  counts2 <- counts
  df <- as_tibble(cbind(counts2,gc=mcols(gr)$gc))
  df <- gather(df, sample, value, -gc)
  ggplot(data=df, aes(x=gc, y=log(value+1)) ) + ylab("log(count + 1)") + xlab("GC content") + geom_hex(bins = 50) + theme_bw() + facet_wrap(~sample, nrow=2) + labs(fill="Number of peaks in hex.")
}

## read in counts
counts <- read.table("../../../data/liu2019_atlas/ATACseq_Matrix/chromatin.accessibility.raw.count.txt", header=TRUE, stringsAsFactors = FALSE)
peakNames <- as.character(counts[,1])
peakNamesSplit <- strsplit(peakNames, split="_")
# remove peaks in random chromosomes
rmPeaks <- which(lapply(peakNamesSplit, length) == 4)
counts <- counts[-rmPeaks,]
peakNamesSplit <- peakNamesSplit[-rmPeaks]
sn <- unlist(lapply(peakNamesSplit, "[[", 1))
sn <- substr(sn, 4, nchar(sn))
start <- as.numeric(unlist(lapply(peakNamesSplit, "[[", 2)))
end <- as.numeric(unlist(lapply(peakNamesSplit, "[[", 3)))
gr <- GRanges(seqnames=sn, ranges = IRanges(start=start, end=end), strand="*")
names(gr) <- paste0("peak", 1:length(gr))

# get GC content: please download genome at ftp://ftp.ensembl.org/pub/release-67/fasta/mus_musculus/dna/
ff <- FaFile("~/data/genomes/mouse/Mus_musculus.NCBIM37.67.dna.toplevel.fa")
peakSeqs <- getSeq(x=ff, gr)
gcContentPeaks <- letterFrequency(peakSeqs, "GC",as.prob=TRUE)[,1]
gcGroups <- Hmisc::cut2(gcContentPeaks, g=20)
mcols(gr)$gc <- gcContentPeaks
hist(gcContentPeaks, xlab="GC content", breaks=40)

# remove peak ranges
counts <- counts[,-1]
# restrict to samples discussed in paper
counts <- as.matrix(counts[,!substr(colnames(counts),1,2) == "D0"])

# metadata
gender <- factor(unlist(lapply(strsplit(colnames(counts), split="_"), "[[", 1)))
tissue <- factor(unlist(lapply(strsplit(colnames(counts), split="_"), function(x){
  paste0(x[2:(length(x)-1)], collapse="_")
})))

# define number of clusters and truth
truth <- droplevels(interaction(gender, tissue))
k <- nlevels(truth)

# explore GC content effect in relation to average accessibility
plotGCHex(gr, rowMeans(counts))
```

# DESeq2

```{r}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(counts,
                              colData=data.frame(tissue, gender),
                              design=as.formula("~ tissue + gender"))
# calculate DESeq2 size factors
dds <- estimateSizeFactors(dds)
# extract the size factors and save them in an object
sf <- sizeFactors(dds)
# divide each row by the vector of size factors
normCountsDESeq2 <- sweep(counts, 2, sizeFactors(dds), "/")
```

# edgeR

```{r}
library(edgeR)
d <- DGEList(counts)
d <- calcNormFactors(d)
nf <- d$samples$norm.factors
effLibSize <- colSums(counts) * nf
effLibSize_scaled <- effLibSize / mean(effLibSize)
# divide each row by the vector of size factors
normCountsEdgeR <- sweep(counts, 2, effLibSize_scaled, "/")
```

# Full quantile normalization

```{r}
normCountsFQ <- FQnorm(counts, type="median")
```

# cqn

```{r}
# run cqn model
cqnObj <- cqn(counts, x=gcContentPeaks, lengths=width(gr))
# calculate normalized counts
normCountsCqn <- 2^(cqnObj$y + cqnObj$offset)

cqnplot(cqnObj, n = 1, xlab = "GC content", ylim = c(-1.5,7), col=tissue)
cqnplot(cqnObj, n = 2, xlab = "Peak width", ylim = c(-1.5,7), col=tissue)
```

# EDASeq

```{r}
library(EDASeq)
wit <- withinLaneNormalization(as.matrix(counts), gcContentPeaks, which="full", num.bins=50)
normCountsEDASeq <- betweenLaneNormalization(wit, which="full")
rm(wit)
```

# GC-QN

```{r}
gcGroups <- Hmisc::cut2(gcContentPeaks, g=50)
normCountsGcqn <- gcqn(counts, gcGroups, "median")
```

## Smooth GC-QN

```{r}
normCountsGcqnSmooth <- gcqn_qsmooth(counts, gcGroups, bio=tissue)
```


# Hierarchical clustering

Here we check whether the hierarchy and partitioning of the dataset according to Euclidean distances makes sense biologically.
We would expect each tissue to be separate, and furthermore that the hierarchy is sensible biologically (e.g., large and small intesting are related).

```{r}
library(dendextend)

ddRaw <- dist(log1p(t(counts)), method="euclidean")
ddEdgeR <- dist(log1p(t(normCountsEdgeR)), method="euclidean")
ddDESeq2 <- dist(log1p(t(normCountsDESeq2)), method="euclidean")
ddFQ <- dist(log1p(t(normCountsFQ)), method="euclidean")
ddCqn <- dist(log1p(t(normCountsCqn)), method="euclidean")
ddEDASeq <- dist(log1p(t(normCountsEDASeq)), method="euclidean")
ddGCQN <- dist(log1p(t(normCountsGcqn)), method="euclidean")
ddGCQNSmooth <- dist(log1p(t(normCountsGcqnSmooth)), method="euclidean")


plotDD <- function(distMat, col, label, main, cex=1/2, ...){
  require(dendextend)
  hdd <- hclust(distMat)
  dhdd <- as.dendrogram(hdd)
  origLabels <- labels(dhdd)
  labels(dhdd) <- label[origLabels]
  labels_colors(dhdd) <- col[origLabels]
  dhdd <- set(dhdd, "labels_cex", cex)
  plot(dhdd, main=main, ...)
}

colors <- c(palette(), brewer.pal(9, "Dark2"), "steelblue", "darkseagreen3", "blue")
colTissue <- colors[as.numeric(tissue)]
names(colTissue) <- colnames(counts)
label <- as.character(tissue)
names(label) <- colnames(counts)
# pdf("~/Dropbox/research/atacseq/bulk/plots/hclust_tissue_liu2019.pdf",
#     width=7, height=5)
plotDD(ddRaw, col=colTissue, label=label, main="Raw counts: tissue effect", cex=3/4)
plotDD(ddEdgeR, col=colTissue, label=label, main="TMM: tissue effect", cex=3/4)
plotDD(ddDESeq2, col=colTissue, label=label, main="DESeq2 MOR: tissue effect", cex=3/4)
plotDD(ddFQ, col=colTissue, label=label, main="FQ: tissue effect", cex=3/4)
plotDD(ddCqn, col=colTissue, label=label, main="cqn: tissue effect", cex=3/4)
plotDD(ddEDASeq, col=colTissue, label=label, main="FQ-FQ: tissue effect", cex=3/4)
plotDD(ddGCQN, col=colTissue, label=label, main="GC-QN: tissue effect", cex=3/4)
plotDD(ddGCQNSmooth, col=colTissue, label=label, main="smooth GC-QN: tissue effect", cex=3/4)
# dev.off()

```

# Differential accessibility

```{r}
library(limma)
library(edgeR)

testVoom <- function(counts, design, L_tissue){
  d <- DGEList(counts)
  v <- voom(d, design)
  fit <- lmFit(v, design)
  fitcon <- contrasts.fit(fit, contrast=L_tissue)
  fitcon <- eBayes(fitcon)
  tt <- topTable(fitcon, number=nrow(counts), sort.by="none")
  return(tt)
}

testEdgeR <- function(counts, design, L_tissue, tmm=FALSE, offset=NULL){
  d <- DGEList(counts)
  if(tmm) d <- calcNormFactors(d)
  if(!is.null(offset)) d$offset <- offset
  d <- estimateDisp(d, design)
  fit <- glmFit(d, design)
  lrt <- glmLRT(fit, contrast=L_tissue)
  lrt$table$padj <- p.adjust(lrt$table$PValue, "fdr")
  return(lrt)
}

covarDf <- data.frame(tissue, gender)
testDESeq2 <- function(counts, covarDf, L_tissue){
  dds <- DESeqDataSetFromMatrix(counts,
                              colData=covarDf,
                              design=as.formula("~ tissue + gender"))
  dds <- estimateSizeFactors(dds)
  dds <- DESeq(dds)
  res <- results(dds, contrast = L_tissue)
  return(res)
}

design <- model.matrix(~tissue+gender)
L_heartLiver <- matrix(0, nrow=ncol(design), ncol=1)
rownames(L_heartLiver) <- colnames(design)
L_heartLiver["tissueHeart",1] <- 1
L_heartLiver["tissueLiver",1] <- -1

resList <- list()
normMethods <- c("Raw", "DESeq2", "EdgeR", "FQ", "EDASeq", "Cqn", "Gcqn", "GcqnSmooth")
# for(mm in 1:length(normMethods)){
#   message(mm)
#   curMethod <- normMethods[mm]
#   if(curMethod == "Raw"){
#     resList[[mm]] <- testEdgeR(counts, design, L_heartLiver)
#   } else if(curMethod == "DESeq2"){
#     resList[[mm]] <- testDESeq2(counts, covarDf, L_heartLiver)
#   } else if(curMethod == "EdgeR"){
#     resList[[mm]] <- testEdgeR(counts, design, L_heartLiver, tmm=TRUE)
#   } else if(curMethod == "Cqn"){
#     resList[[mm]] <- testEdgeR(counts, design, L_heartLiver, offset=cqnObj$glm.offset)
#   } else if(curMethod %in% c("FQ", "EDASeq", "Gcqn", "GcqnSmooth")){
#     curCounts <- get(paste0("normCounts", curMethod))
#     resList[[mm]] <- testEdgeR(curCounts, design, L_heartLiver)
#   }
# }
# saveRDS(resList, file="resList.rds")
```

## Visualization

```{r}
resList <- readRDS("resList.rds")
normMethods[which(normMethods == "EDASeq")] <- "FQ-FQ"
normMethods[which(normMethods == "Gcqn")] <- "GC-FQ"
normMethods[which(normMethods == "GcqnSmooth")] <- "smooth GC-FQ"
normMethods[which(normMethods == "RUV")] <- "RUVSeq"
names(resList) <- normMethods
resList <- resList[-length(resList)]

### visualization
# fold change bias
# png("~/Dropbox/research/atacseq/bulk/plots/liu_foldChange.png", width=8, units="in", height=7, res=200)
mypar(mfrow=c(3,3), mar=c(4,4,2,1), bty='l', las=2, cex.axis=3/2, cex.lab=3/2, cex.main=2)
for(ii in 1:length(resList)){
  tab <- resList[[ii]]
  if(ii == 2){
    boxplot(tab$log2FoldChange ~ gcGroups, ylim=c(2,-2), main=names(resList)[ii],
            xaxt="n", xlab="GC-content bin", ylab="Log2 fold change")
    abline(h=0, lwd=2, col="red", lty=2)
  } else {
    boxplot(tab$table$logFC ~ gcGroups, ylim=c(2,-2), main=names(resList)[ii],
            xaxt="n", xlab="GC-content bin", ylab="Log2 fold change")
    abline(h=0, lwd=2, col="red", lty=2)
  }
}
# dev.off()

### check distribution of GC-content for top 1000 peaks.
# pdf("~/Dropbox/research/atacseq/bulk/plots/liu_topPeaks.pdf", width=8)
mypar(mfrow=c(3,3))
for(ii in 1:length(resList)){
  tab <- resList[[ii]]
  if(ii == 2){
    ## DESeq2
    topPeaks <- order(tab$stat, decreasing=TRUE)[1:5e3]
  } else {
    ## edgeR
    topPeaks <- order(tab$table$LR, decreasing=TRUE)[1:5e3]
  }
  barplot(table(gcGroups[topPeaks]), ylim=c(0,250), main=names(resList)[ii], xaxt="n", xlab="GC-content bin", ylab="Nr. of peaks")
}
# dev.off()

# all significant peaks
nrPeaks <- c()
mypar(mfrow=c(3,3))
for(ii in 1:length(resList)){
  tab <- resList[[ii]]
  if(ii == 2){
    ## DESeq2
    sigPeaks <- which(tab$padj <= 0.05)
  } else {
    ## edgeR
    sigPeaks <- which(p.adjust(tab$table$PValue, "fdr") <= 0.05)
  }
  barplot(table(gcGroups[sigPeaks]), ylim=c(0,3200), main=names(resList)[ii], xaxt="n", xlab="GC-content bin")
  nrPeaks[ii] <- length(sigPeaks)
}
nrPeaks

# all significant peaks: relative
# pdf("~/Dropbox/research/atacseq/bulk/plots/liu_relnrPeaks.pdf", width=8)
mypar(mfrow=c(3,3), mar=c(4,4,2,1), bty='l', las=2, cex.axis=3/2, cex.lab=3/2, cex.main=2)
for(ii in 1:length(resList)){
  tab <- resList[[ii]]
  if(ii == 2){
    ## DESeq2
    sigPeaks <- which(tab$padj <= 0.05)
  } else {
    ## edgeR
    sigPeaks <- which(p.adjust(tab$table$PValue, "fdr") <= 0.05)
  }
  
  barplot(table(gcGroups[sigPeaks]) / table(gcGroups[sigPeaks])[1], ylim=c(0,4), main=names(resList)[ii],xaxt="n", yaxt='n', xlab="GC-content bin")
  axis(side=2, at=c(0,1,2,3))
  title(ylab="Relative # peaks", line=1.3, cex.lab=3/2)
  abline(h=1, col="red", lwd=2, lty=2)
}
# dev.off()

names(nrPeaks) <- names(resList)
barplot(nrPeaks, las=2)
```




