---
title: "GC-content effects across technical replicates"
author: "Koen Van den Berge"
date: "8/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read in peak counts and get GC content

```{r}
library(GenomicAlignments)
library(ggplot2)
library(edgeR)

df <- read.csv("../../data/geschwind2018_GSE95023/GSE95023_readswithinpeaks.csv.gz")
counts <- df[,-c(1:3)]
gr <- GRanges(seqnames=df$CHR, ranges=IRanges(df$START, df$END), strand="*", mcols=counts)
ff <- FaFile("~/data/genomes/human/Homo_sapiens.GRCh37.75.dna_sm.primary_assembly.fa.gz")
peakSeqs <- getSeq(x=ff, gr)
gcContentPeaks <- letterFrequency(peakSeqs, "GC",as.prob=TRUE)[,1]
gcGroups <- Hmisc::cut2(gcContentPeaks, g=20)
```

# Design

Samples are derived from three individual donors.
For every donor, neural progenitor (GZ) zone and neuron rich region (CP) is sequenced.
For every sample, three to four technical replicates are available.

```{r}
patient <- factor(c(rep("A",7), rep("B",6), rep("C",6)))
region <- factor(c(rep("GZ",4), rep("CP",3), rep("GZ",3), rep("CP",3), rep("GZ",3), rep("CP",3)))

table(patient,region)
```


```{r}
lowListGC <- list()
for(kk in 1:ncol(counts)){
  set.seed(kk)
  lowListGC[[kk]] <- lowess(x=gcContentPeaks, y=log1p(counts[,kk]), f=1/10)
}
```



```{r}
combs <- expand.grid(unique(patient),unique(region))
colnames(combs) <- c("patient", "region")

plist <- list()
for(cc in 1:nrow(combs)){
  lowCur <- lowListGC[patient == combs[cc, "patient"] & region == combs[cc, "region"]]
  
  dfList <- list()
  for(ss in 1:length(lowCur)){
    oox <- order(lowCur[[ss]]$x)
    dfList[[ss]] <- data.frame(x=lowCur[[ss]]$x[oox], y=lowCur[[ss]]$y[oox], sample=ss)
  }
  dfAll <- do.call(rbind, dfList)
  dfAll$sample <- factor(dfAll$sample)
  
  ## just the average GC content
  pCur <- ggplot(dfAll, aes(x=x, y=y, group=sample, color=sample)) +
    geom_line(size = 1) +
    xlab("GC-content") +
    ylab("log(count + 1)") +
    theme_classic()

  plist[[cc]] <- pCur
}

cowplot::plot_grid(plotlist = plist)
```

# Mock comparison 

There is one sample with 4 technical replicates, namely patient A and region GZ. We will perform a mock comparison to assess GC-content effects.


```{r}
subCounts <- counts[,patient == "A" & region == "GZ"]

mock <- c(2,2,1,1)
design <- model.matrix(~mock)

d <- DGEList(subCounts)
d <- calcNormFactors(d)
d <- estimateDisp(d, design)
fit <- glmFit(d, design)
lrt <- glmLRT(fit, coef=2)

par(bty='l')
boxplot(lrt$table$logFC ~ gcGroups,
        ylim=c(-1,1), ylab="logFC", xlab="GC-content bin (low to high)",
        xaxt="n")
abline(h=0, col="red")

dfEdgeR <- data.frame(logFC=log(2^lrt$table$logFC),
                 gc=gcGroups)
pedgeR <- ggplot(dfEdgeR) +
  aes(x=gc, y=logFC, color=gc) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  scale_color_manual(values=wesanderson::wes_palette("Zissou1", nlevels(gcGroups), "continuous")) +
  geom_abline(intercept = 0, slope = 0, col="black", lty=2) +
  theme_bw() + 
  ylim(c(-1,1)) +
  ggtitle("TMM normalization") +
  xlab("GC-content bin") +
  theme(axis.text.x = element_text(angle = 45, vjust = .5),
        legend.position = "none",
        axis.title = element_text(size=16)) +
  geom_smooth(se=FALSE, color="blue", aes(group=1), lwd=1)
pedgeR

```


# Lowess fits

```{r}
colsBig <- clusterExperiment:::massivePalette

regionPatient <- factor(paste0(region, patient))

lowListGC <- list()
for(kk in 1:ncol(counts)){
  set.seed(kk)
  lowListGC[[kk]] <- lowess(x=gcContentPeaks, y=log1p(counts[,kk]), f=1/10)
}

for(cc in 1:nlevels(regionPatient)){
  curCT <- levels(regionPatient)[cc]
  id <- which(regionPatient == curCT)
  plot(x=seq(min(gcContentPeaks), max(gcContentPeaks), length=10),
     y=seq(2.8, 5, length=10), type='n',
     xlab="GC-content", ylab="log(count + 1)", main=curCT)
  for(ii in 1:length(id)){
    curID <- id[ii]
    oo <- order(lowListGC[[curID]]$x)
    lines(x=lowListGC[[curID]]$x[oo], y=lowListGC[[curID]]$y[oo], col=colsBig)
  }
}

# across all cell types
set.seed(44)
pList <- c()
id <- sample(nrow(counts), size=1e4)
for(cc in 1:nlevels(regionPatient)){
  curCT <- levels(regionPatient)[cc]
  lowCT <- lowListGC[regionPatient == curCT]
  dfList <- list()
  for(ss in 1:length(lowCT)){
  oox <- order(lowCT[[ss]]$x[id])
  dfList[[ss]] <- data.frame(x=lowCT[[ss]]$x[id][oox], y=lowCT[[ss]]$y[id][oox], sample=ss)
  }
  dfAll <- do.call(rbind, dfList)
  dfAll$sample <- factor(dfAll$sample)
  pCT <- ggplot(dfAll, aes(x=x, y=y, group=sample, color=sample)) +
    geom_line(size = 1) +
    xlab("GC-content") +
    ylab("log(count + 1)") +
    theme_classic() +
    ggtitle(curCT) +
    theme(legend.position = "none") +
    ylim(c(2.8, 5.5))
  pList[[cc]] <- pCT
}

cowplot::plot_grid(plotlist=pList, nrow=2, ncol=3)

```

