---
title: "GC-content effects across technical replicates"
author: "Koen Van den Berge"
date: "8/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read in peak counts and get GC content

```{r}
library(GenomicAlignments)
library(ggplot2)
library(edgeR)

df <- read.csv("../../data/geschwind2018_GSE95023/GSE95023_readswithinpeaks.csv.gz")
counts <- df[,-c(1:3)]
gr <- GRanges(seqnames=df$CHR, ranges=IRanges(df$START, df$END), strand="*", mcols=counts)
ff <- FaFile("~/data/genomes/human/Homo_sapiens.GRCh37.75.dna_sm.primary_assembly.fa.gz")
peakSeqs <- getSeq(x=ff, gr)
gcContentPeaks <- letterFrequency(peakSeqs, "GC",as.prob=TRUE)[,1]
gcGroups <- Hmisc::cut2(gcContentPeaks, g=20)
```

# Design

Samples are derived from three individual donors.
For every donor, neural progenitor (GZ) zone and neuron rich region (CP) is sequenced.
For every sample, three to four technical replicates are available.

```{r}
patient <- factor(c(rep("A",7), rep("B",6), rep("C",6)))
region <- factor(c(rep("GZ",4), rep("CP",3), rep("GZ",3), rep("CP",3), rep("GZ",3), rep("CP",3)))

table(patient,region)
```


```{r}
lowListGC <- list()
for(kk in 1:ncol(counts)){
  set.seed(kk)
  lowListGC[[kk]] <- lowess(x=gcContentPeaks, y=log1p(counts[,kk]), f=1/10)
}
```



```{r}
combs <- expand.grid(unique(patient),unique(region))
colnames(combs) <- c("patient", "region")

plist <- list()
for(cc in 1:nrow(combs)){
  lowCur <- lowListGC[patient == combs[cc, "patient"] & region == combs[cc, "region"]]
  
  dfList <- list()
  for(ss in 1:length(lowCur)){
    oox <- order(lowCur[[ss]]$x)
    dfList[[ss]] <- data.frame(x=lowCur[[ss]]$x[oox], y=lowCur[[ss]]$y[oox], sample=ss)
  }
  dfAll <- do.call(rbind, dfList)
  dfAll$sample <- factor(dfAll$sample)
  
  ## just the average GC content
  pCur <- ggplot(dfAll, aes(x=x, y=y, group=sample, color=sample)) +
    geom_line(size = 1) +
    xlab("GC-content") +
    ylab("log(count + 1)") +
    theme_classic()

  plist[[cc]] <- pCur
}

cowplot::plot_grid(plotlist = plist)
```

# Mock comparison 

There is one sample with 4 technical replicates, namely patient A and region GZ. We will perform a mock comparison to assess GC-content effects.


```{r}
subCounts <- counts[,patient == "A" & region == "GZ"]

mock <- c(2,2,1,1)
design <- model.matrix(~mock)

d <- DGEList(subCounts)
d <- calcNormFactors(d)
d <- estimateDisp(d, design)
fit <- glmFit(d, design)
lrt <- glmLRT(fit, coef=2)

par(bty='l')
boxplot(lrt$table$logFC ~ gcGroups,
        ylim=c(-1,1), ylab="logFC", xlab="GC-content bin (low to high)",
        xaxt="n")
abline(h=0, col="red")

dfEdgeR <- data.frame(logFC=log(2^lrt$table$logFC),
                 gc=gcGroups)
pedgeR <- ggplot(dfEdgeR) +
  aes(x=gc, y=logFC, color=gc) +
  geom_violin() +
  geom_boxplot(width=0.1) +
  scale_color_manual(values=wesanderson::wes_palette("Zissou1", nlevels(gcGroups), "continuous")) +
  geom_abline(intercept = 0, slope = 0, col="black", lty=2) +
  theme_bw() + 
  ylim(c(-1,1)) +
  ggtitle("TMM normalization") +
  xlab("GC-content bin") +
  theme(axis.text.x = element_text(angle = 45, vjust = .5),
        legend.position = "none",
        axis.title = element_text(size=16)) +
  geom_smooth(se=FALSE, color="blue", aes(group=1), lwd=1)
pedgeR

```

# Biological comparison

Below, we compare the GZ and CP regions for patient A, the patient for whom we had also performed the mock comparison.

```{r}
subCounts <- counts[,patient == "A"]
regionA <- region[patient == "A"]
design <- model.matrix(~regionA)

d <- DGEList(subCounts)
d <- calcNormFactors(d)
d <- estimateDisp(d, design)
fit <- glmFit(d, design)
lrtBio <- glmLRT(fit, coef=2)

par(bty='l')
boxplot(lrtBio$table$logFC ~ gcGroups,
        ylim=c(-1,1), ylab="logFC", xlab="GC-content bin (low to high)",
        xaxt="n")
abline(h=0, col="red")
```

# All mock comparisons: Is bias technical or biological?

Perform all within-group comparisons and contrast with the between-group comparison.

```{r}
subCountsGZ <- counts[,patient == "A" & region == "GZ"]
subsetPossibilities <- combn(1:4,2)
lrtTableListGZ <- list()
for(ii in 1:ncol(subsetPossibilities)){
  message(ii)
  groups <- rep("A",4)
  groups[subsetPossibilities[,ii]] <- "B"
  curDesign <- model.matrix(~groups)
  dGZ <- DGEList(subCountsGZ)
  dGZ <- calcNormFactors(dGZ)
  dGZ <- estimateDisp(dGZ, curDesign)
  fitGZ <- glmFit(dGZ, curDesign)
  lrtGZ <- glmLRT(fitGZ, coef=2)
  lrtTableListGZ[[ii]] <- lrtGZ$table
}


subCountsCP <- counts[,patient == "A" & region == "CP"]
subsetPossibilities <- combn(1:3,1)
lrtTableListCP <- list()
for(ii in 1:ncol(subsetPossibilities)){
  message(ii)
  groups <- rep("A",3)
  groups[subsetPossibilities[,ii]] <- "B"
  curDesign <- model.matrix(~groups)
  dCP <- DGEList(subCountsCP)
  dCP <- calcNormFactors(dCP)
  dCP <- estimateDisp(dCP, curDesign)
  fitCP <- glmFit(dCP, curDesign)
  lrtCP <- glmLRT(fitCP, coef=2)
  lrtTableListCP[[ii]] <- lrtCP$table
}

## within group GZ
rafalib::mypar(mfrow=c(3,2))
for(ii in 1:length(lrtTableListGZ)){
  par(bty='l')
  boxplot(lrtTableListGZ[[ii]]$logFC ~ gcGroups,
          ylim=c(-1,1), ylab="logFC", xlab="GC-content bin (low to high)",
          xaxt="n")
  abline(h=0, col="red")
}

## within group CP
rafalib::mypar(mfrow=c(2,2))
for(ii in 1:length(lrtTableListCP)){
  par(bty='l')
  boxplot(lrtTableListCP[[ii]]$logFC ~ gcGroups,
          ylim=c(-1,1), ylab="logFC", xlab="GC-content bin (low to high)",
          xaxt="n")
  abline(h=0, col="red")
}

## between groups
rafalib::mypar(mfrow=c(1,1))
par(bty='l')
boxplot(lrtBio$table$logFC ~ gcGroups,
        ylim=c(-1,1), ylab="logFC", xlab="GC-content bin (low to high)",
        xaxt="n", main="Between-group")
abline(h=0, col="red")

```

## All on one figure

```{r}

for(ii in 1:length(lrtTableListGZ)){
  par(bty='l')
  curLowess <- lowess(x=gcContentPeaks, y=lrtTableListGZ[[ii]]$logFC, f=1/5)
  oo <- order(curLowess$x)
  if(ii == 1){
    plot(x=curLowess$x[oo], y=curLowess$y[oo], type='l', ylim=c(-1/2, 1/2),
         xlab = "GC-content", ylab = "Log-fold-change", col="gray")
  } else {
    lines(x=curLowess$x[oo], y=curLowess$y[oo], col="gray")
  }
}

for(ii in 1:length(lrtTableListCP)){
  par(bty='l')
  curLowess <- lowess(x=gcContentPeaks, y=lrtTableListCP[[ii]]$logFC, f=1/5)
  oo <- order(curLowess$x)
  lines(x=curLowess$x[oo], y=curLowess$y[oo], col="gray")
}


lowessBio <- lowess(x=gcContentPeaks, y=lrtBio$table$logFC, f=1/10)
oo <- order(lowessBio$x)
lines(x=lowessBio$x[oo], y=lowessBio$y[oo], col="steelblue", lwd=2)
abline(h=0, col="red")
legend("topleft", c("Technical comparison", "Biological comparison"),
       lty=1, col=c("grey", "steelblue"), bty='n', cex=1, lwd=2)

```

# Save objects

```{r}
saveRDS(lrtTableListGZ, file="../../objects/technicalGC/torreUbieta_lrtTableListGZ.rds")
saveRDS(lrtTableListCP, file="../../objects/technicalGC/torreUbieta_lrtTableListCP.rds")
saveRDS(lrtBio, file="../../objects/technicalGC/torreUbieta_lrtBio.rds")
saveRDS(gcContentPeaks, file="../../objects/technicalGC/torreUbieta_gcContentPeaks.rds")

```




# Lowess fits

```{r}
colsBig <- clusterExperiment:::massivePalette

regionPatient <- factor(paste0(region, patient))

lowListGC <- list()
for(kk in 1:ncol(counts)){
  set.seed(kk)
  lowListGC[[kk]] <- lowess(x=gcContentPeaks, y=log1p(counts[,kk]), f=1/10)
}

for(cc in 1:nlevels(regionPatient)){
  curCT <- levels(regionPatient)[cc]
  id <- which(regionPatient == curCT)
  plot(x=seq(min(gcContentPeaks), max(gcContentPeaks), length=10),
     y=seq(2.8, 5, length=10), type='n',
     xlab="GC-content", ylab="log(count + 1)", main=curCT)
  for(ii in 1:length(id)){
    curID <- id[ii]
    oo <- order(lowListGC[[curID]]$x)
    lines(x=lowListGC[[curID]]$x[oo], y=lowListGC[[curID]]$y[oo], col=colsBig)
  }
}

# across all cell types
set.seed(44)
pList <- c()
id <- sample(nrow(counts), size=1e4)
for(cc in 1:nlevels(regionPatient)){
  curCT <- levels(regionPatient)[cc]
  lowCT <- lowListGC[regionPatient == curCT]
  dfList <- list()
  for(ss in 1:length(lowCT)){
  oox <- order(lowCT[[ss]]$x[id])
  dfList[[ss]] <- data.frame(x=lowCT[[ss]]$x[id][oox], y=lowCT[[ss]]$y[id][oox], sample=ss)
  }
  dfAll <- do.call(rbind, dfList)
  dfAll$sample <- factor(dfAll$sample)
  pCT <- ggplot(dfAll, aes(x=x, y=y, group=sample, color=sample)) +
    geom_line(size = 1) +
    xlab("GC-content") +
    ylab("log(count + 1)") +
    theme_classic() +
    ggtitle(curCT) +
    theme(legend.position = "none") +
    ylim(c(2.8, 5.5))
  pList[[cc]] <- pCT
}

cowplot::plot_grid(plotlist=pList, nrow=2, ncol=3)

```

