---
title: "ATAC-seq normalization summary, polished"
author: "Koen Van den Berge"
date: "12/12/2019"
output: html_document
---


```{r}
getMeanScoreRank <- function(scores){
  scoreT <- t(scores)
  ranked_scores <- apply(scoreT,1,rank)
  mean_score_rank <- rowMeans(ranked_scores)
  mean_score_rank <- mean_score_rank[order(mean_score_rank, decreasing=TRUE)]
  return(mean_score_rank)
}

rmMethods <- c("gcqn_mean", "gcqn_median_10", "gcqn_median_20",  "gcqn_median_100",
               "gcqn_median_permuted")

library(ggplot2)
filesDir <- "../../objects/"
files <- list.files(filesDir,
                    pattern = "*scores.rds$")
datasets <- unlist(lapply(strsplit(files, split="_"),"[[",1))
datasets[datasets %in% "geschwind2018"] <- "torre-ubieta2018"
scores <- sapply(1:length(files), function(ii) readRDS(paste0(filesDir,files[ii])))
# remove other methods
scores <- lapply(scores, function(scoreMat){
  scoreNames <- rownames(scoreMat)
  rmID <- unlist(sapply(rmMethods, function(method){
    grep(x=scoreNames, pattern=method)
  }))
  return(scoreMat[-rmID,])
})
scores <- lapply(scores, function(x) x[,!colnames(x) %in% c("RLE_MED","RLE_IQR")])
scoreRanks <- lapply(scores, getMeanScoreRank)
rankVector <- do.call(c,scoreRanks)
scoreDfLong <- data.frame(method=names(rankVector),
                          dataset=unlist(mapply(rep,datasets,unlist(lapply(scoreRanks,length)))),
                          rank=rankVector)


pal <- wesanderson::wes_palette("Zissou1", n=40, type="continuous")
ggplot(scoreDfLong, aes(y=method, x=dataset, fill=rank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Scone rank')
```


# Order by average rank

```{r}
library(dplyr)
rankAv <- scoreDfLong %>% group_by(method) %>% summarize(avgRank=mean(rank, na.rm=TRUE))
oo <- order(rankAv$avgRank, decreasing=TRUE)
orderedMethods <- rankAv$method[oo]


scoreDfOrdered <- scoreDfLong
scoreDfOrdered$method <- factor(scoreDfOrdered$method, levels=orderedMethods)
ggplot(scoreDfOrdered, aes(y=method, x=dataset, fill=rank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Scone rank')
```


# Conditional on the dimension of unwanted variation latent variables: what performs best?

Here we stratify performances based on the number of estimated unwanted variation variables that was adjusted for after the normalization.

Since different methods have different domains in terms of the ranking, we scale the ranks within each dataset before taking the average.

```{r}
gcNormMethods <- c("FQ-FQ", "GC-FQ", "cqn_length", "cqn", "GC-FQ_smooth")
```


## No latent variables

```{r}
id0 <- grep(x=scoreDfLong$method, pattern="no_uv")
scoreDf0 <- scoreDfLong[id0,]
scoreDf0$methodShort <- unlist(lapply(strsplit(as.character(scoreDf0$method), split=","), "[[", 2))
scoreDf0$methodShort[scoreDf0$methodShort == "gcqn_median_50"] <- "GC-FQ"
scoreDf0$methodShort[scoreDf0$methodShort == "edaseq"] <- "FQ-FQ"
scoreDf0$methodShort[scoreDf0$methodShort == "gcqn_smooth"] <- "GC-FQ_smooth"
scoreDf0$methodShort[scoreDf0$methodShort == "fq"] <- "FQ"
scoreDf0$methodShort[scoreDf0$methodShort == "tmm"] <- "TMM"
scoreDf0$methodShort[scoreDf0$methodShort == "deseq"] <- "DESeq2"
scoreDf0$methodShort[scoreDf0$methodShort == "uq"] <- "UQ"
scoreDf0$methodShort[scoreDf0$methodShort == "none"] <- "None"
scoreDf0$methodShort[scoreDf0$methodShort == "sum"] <- "Sum"
scoreDf0$methodShort <- factor(scoreDf0$methodShort)

## scale ranks
scoreDf0 <- scoreDf0 %>% group_by(dataset) %>% mutate(scaledRank=scale(rank))

## take average of scaled ranks
avRank0 <- scoreDf0 %>% group_by(methodShort) %>% summarize(avRank=mean(scaledRank))
oo0 <- order(avRank0$avRank, decreasing=TRUE)
orderedMethodShort0 <- avRank0$methodShort[oo0]

scoreDfOrdered0 <- scoreDf0
scoreDfOrdered0$methodShort <- factor(scoreDfOrdered0$methodShort, levels=orderedMethodShort0)
ggplot(scoreDfOrdered0, aes(y=methodShort, x=dataset, fill=scaledRank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Scaled Scone\nrank') +
  ggtitle("Without RUV variables") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9)) + 
  theme(panel.background = element_rect(fill="white")) + 
  theme(axis.text.y=element_text(colour=ifelse(orderedMethodShort0 %in% gcNormMethods, "dodgerblue", "black")))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_noRUV_noGCRLE.pdf")
```

## One latent variable

```{r}
id1 <- grep(x=scoreDfLong$method, pattern="ruv_k=1")
scoreDf1 <- scoreDfLong[id1,]
scoreDf1$methodShort <- unlist(lapply(strsplit(as.character(scoreDf1$method), split=","), "[[", 2))
scoreDf1$methodShort[scoreDf1$methodShort == "gcqn_median_50"] <- "GC-FQ"
scoreDf1$methodShort[scoreDf1$methodShort == "edaseq"] <- "FQ-FQ"
scoreDf1$methodShort[scoreDf1$methodShort == "gcqn_smooth"] <- "GC-FQ_smooth"
scoreDf1$methodShort[scoreDf1$methodShort == "fq"] <- "FQ"
scoreDf1$methodShort[scoreDf1$methodShort == "tmm"] <- "TMM"
scoreDf1$methodShort[scoreDf1$methodShort == "deseq"] <- "DESeq2"
scoreDf1$methodShort[scoreDf1$methodShort == "uq"] <- "UQ"
scoreDf1$methodShort[scoreDf1$methodShort == "none"] <- "None"
scoreDf1$methodShort[scoreDf1$methodShort == "sum"] <- "Sum"
scoreDf1$methodShort <- factor(scoreDf1$methodShort)

## scale ranks
scoreDf1 <- scoreDf1 %>% group_by(dataset) %>% mutate(scaledRank=scale(rank))

avRank1 <- scoreDf1 %>% group_by(methodShort) %>% summarize(avRank=mean(scaledRank))
oo1 <- order(avRank1$avRank, decreasing=TRUE)
orderedMethodShort1 <- avRank1$methodShort[oo1]

scoreDfOrdered1 <- scoreDf1
scoreDfOrdered1$methodShort <- factor(scoreDfOrdered1$methodShort, levels=orderedMethodShort1)
ggplot(scoreDfOrdered1, aes(y=methodShort, x=dataset, fill=scaledRank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Scaled Scone\nrank') +
  ggtitle("One RUV variable") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9))  + 
  theme(panel.background = element_rect(fill="white")) + 
  theme(axis.text.y=element_text(colour=ifelse(orderedMethodShort1 %in% gcNormMethods, "dodgerblue", "black")))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_RUV1_noGCRLE.pdf")
```

## Two latent variables

```{r}
id2 <- grep(x=scoreDfLong$method, pattern="ruv_k=2")
scoreDf2 <- scoreDfLong[id2,]
scoreDf2$methodShort <- unlist(lapply(strsplit(as.character(scoreDf2$method), split=","), "[[", 2))
scoreDf2$methodShort[scoreDf2$methodShort == "gcqn_median_50"] <- "GC-FQ"
scoreDf2$methodShort[scoreDf2$methodShort == "edaseq"] <- "FQ-FQ"
scoreDf2$methodShort[scoreDf2$methodShort == "gcqn_smooth"] <- "GC-FQ_smooth"
scoreDf2$methodShort[scoreDf2$methodShort == "fq"] <- "FQ"
scoreDf2$methodShort[scoreDf2$methodShort == "tmm"] <- "TMM"
scoreDf2$methodShort[scoreDf2$methodShort == "deseq"] <- "DESeq2"
scoreDf2$methodShort[scoreDf2$methodShort == "uq"] <- "UQ"
scoreDf2$methodShort[scoreDf2$methodShort == "none"] <- "None"
scoreDf2$methodShort[scoreDf2$methodShort == "sum"] <- "Sum"
scoreDf2$methodShort <- factor(scoreDf2$methodShort)

## scale ranks
scoreDf2 <- scoreDf2 %>% group_by(dataset) %>% mutate(scaledRank=scale(rank))

avRank2 <- scoreDf2 %>% group_by(methodShort) %>% summarize(avRank=mean(scaledRank))
oo2 <- order(avRank2$avRank, decreasing=TRUE)
orderedMethodShort2 <- avRank2$methodShort[oo2]

scoreDfOrdered2 <- scoreDf2
scoreDfOrdered2$methodShort <- factor(scoreDfOrdered2$methodShort, levels=orderedMethodShort2)
ggplot(scoreDfOrdered2, aes(y=methodShort, x=dataset, fill=scaledRank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Scaled Scone\nrank') +
  ggtitle("Two RUV variables") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9)) + 
  theme(panel.background = element_rect(fill="white")) + 
  theme(axis.text.y=element_text(colour=ifelse(orderedMethodShort2 %in% gcNormMethods, "dodgerblue", "black")))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_RUV2_noGCRLE.pdf")
```

## Three latent variables

```{r}
id3 <- grep(x=scoreDfLong$method, pattern="ruv_k=3")
scoreDf3 <- scoreDfLong[id3,]
scoreDf3$methodShort <- unlist(lapply(strsplit(as.character(scoreDf3$method), split=","), "[[", 2))
scoreDf3$methodShort[scoreDf3$methodShort == "gcqn_median_50"] <- "GC-FQ"
scoreDf3$methodShort[scoreDf3$methodShort == "edaseq"] <- "FQ-FQ"
scoreDf3$methodShort[scoreDf3$methodShort == "gcqn_smooth"] <- "GC-FQ_smooth"
scoreDf3$methodShort[scoreDf3$methodShort == "fq"] <- "FQ"
scoreDf3$methodShort[scoreDf3$methodShort == "tmm"] <- "TMM"
scoreDf3$methodShort[scoreDf3$methodShort == "deseq"] <- "DESeq2"
scoreDf3$methodShort[scoreDf3$methodShort == "uq"] <- "UQ"
scoreDf3$methodShort[scoreDf3$methodShort == "none"] <- "None"
scoreDf3$methodShort[scoreDf3$methodShort == "sum"] <- "Sum"
scoreDf3$methodShort <- factor(scoreDf3$methodShort)

## scale ranks
scoreDf3 <- scoreDf3 %>% group_by(dataset) %>% mutate(scaledRank=scale(rank))

avRank3 <- scoreDf3 %>% group_by(methodShort) %>% summarize(avRank=mean(scaledRank))
oo3 <- order(avRank3$avRank, decreasing=TRUE)
orderedMethodShort3 <- avRank3$methodShort[oo3]

scoreDfOrdered3 <- scoreDf3
scoreDfOrdered3$methodShort <- factor(scoreDfOrdered3$methodShort, levels=orderedMethodShort3)
ggplot(scoreDfOrdered3, aes(y=methodShort, x=dataset, fill=scaledRank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Scaled Scone\nrank') +
  ggtitle("Three RUV variables") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9)) + 
  theme(panel.background = element_rect(fill="white")) + 
  theme(axis.text.y=element_text(colour=ifelse(orderedMethodShort3 %in% gcNormMethods, "dodgerblue", "black")))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_RUV3_noGCRLE.pdf")

```

## Four latent variables

```{r}
id4 <- grep(x=scoreDfLong$method, pattern="ruv_k=4")
scoreDf4 <- scoreDfLong[id4,]
scoreDf4$methodShort <- unlist(lapply(strsplit(as.character(scoreDf4$method), split=","), "[[", 2))
scoreDf4$methodShort[scoreDf4$methodShort == "gcqn_median_50"] <- "GC-FQ"
scoreDf4$methodShort[scoreDf4$methodShort == "edaseq"] <- "FQ-FQ"
scoreDf4$methodShort[scoreDf4$methodShort == "gcqn_smooth"] <- "GC-FQ_smooth"
scoreDf4$methodShort[scoreDf4$methodShort == "fq"] <- "FQ"
scoreDf4$methodShort[scoreDf4$methodShort == "tmm"] <- "TMM"
scoreDf4$methodShort[scoreDf4$methodShort == "deseq"] <- "DESeq2"
scoreDf4$methodShort[scoreDf4$methodShort == "uq"] <- "UQ"
scoreDf4$methodShort[scoreDf4$methodShort == "none"] <- "None"
scoreDf4$methodShort[scoreDf4$methodShort == "sum"] <- "Sum"
scoreDf4$methodShort <- factor(scoreDf4$methodShort)

## scale ranks
scoreDf4 <- scoreDf4 %>% group_by(dataset) %>% mutate(scaledRank=scale(rank))

avRank4 <- scoreDf4 %>% group_by(methodShort) %>% summarize(avRank=mean(scaledRank))
oo4 <- order(avRank4$avRank, decreasing=TRUE)
orderedMethodShort4 <- avRank4$methodShort[oo4]

scoreDfOrdered4 <- scoreDf4
scoreDfOrdered4$methodShort <- factor(scoreDfOrdered4$methodShort, levels=orderedMethodShort4)
ggplot(scoreDfOrdered4, aes(y=methodShort, x=dataset, fill=scaledRank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Scaled Scone\nrank') +
  ggtitle("Four RUV variables") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9)) + 
  theme(panel.background = element_rect(fill="white")) + 
  theme(axis.text.y=element_text(colour=ifelse(orderedMethodShort4 %in% gcNormMethods, "dodgerblue", "black")))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_RUV4_noGCRLE.pdf")

```

## Five latent variables

```{r}
id5 <- grep(x=scoreDfLong$method, pattern="ruv_k=5")
scoreDf5 <- scoreDfLong[id5,]
scoreDf5$methodShort <- unlist(lapply(strsplit(as.character(scoreDf5$method), split=","), "[[", 2))
scoreDf5$methodShort[scoreDf5$methodShort == "gcqn_median_50"] <- "GC-FQ"
scoreDf5$methodShort[scoreDf5$methodShort == "edaseq"] <- "FQ-FQ"
scoreDf5$methodShort[scoreDf5$methodShort == "gcqn_smooth"] <- "GC-FQ_smooth"
scoreDf5$methodShort[scoreDf5$methodShort == "fq"] <- "FQ"
scoreDf5$methodShort[scoreDf5$methodShort == "tmm"] <- "TMM"
scoreDf5$methodShort[scoreDf5$methodShort == "deseq"] <- "DESeq2"
scoreDf5$methodShort[scoreDf5$methodShort == "uq"] <- "UQ"
scoreDf5$methodShort[scoreDf5$methodShort == "none"] <- "None"
scoreDf5$methodShort[scoreDf5$methodShort == "sum"] <- "Sum"
scoreDf5$methodShort <- factor(scoreDf5$methodShort)

## scale ranks
scoreDf5 <- scoreDf5 %>% group_by(dataset) %>% mutate(scaledRank=scale(rank))

avRank5 <- scoreDf5 %>% group_by(methodShort) %>% summarize(avRank=mean(scaledRank))
oo5 <- order(avRank5$avRank, decreasing=TRUE)
orderedMethodShort5 <- avRank5$methodShort[oo5]

scoreDfOrdered5 <- scoreDf5
scoreDfOrdered5$methodShort <- factor(scoreDfOrdered5$methodShort, levels=orderedMethodShort5)
ggplot(scoreDfOrdered5, aes(y=methodShort, x=dataset, fill=scaledRank)) + 
  geom_tile() + 
  scale_fill_gradientn(colours = pal, name = 'Scaled Scone\nrank') +
  ggtitle("Five RUV variables") +
  ylab("") + xlab("Dataset") +
  theme(axis.text.x = element_text(angle = 30, vjust= 0.7, size=9),
        axis.text.y = element_text(size=9)) + 
  theme(panel.background = element_rect(fill="white")) + 
  theme(axis.text.y=element_text(colour=ifelse(orderedMethodShort5 %in% gcNormMethods, "dodgerblue", "black")))
ggsave("~/Dropbox/research/atacseq/bulk/plots/benchmark_RUV5_noGCRLE.pdf")
```




