---
title: "Total benchmark summary"
author: "Koen Van den Berge"
date: "3/16/2021"
output: 
  html_document:
    toc: true
    toc_float: true
---


We have the following evaluation metrics for a scone evaluation:

 - BIO_SIL
 - BATCH_SIL
 - PAM_SIL
 - EXP_QC_COR
 - EXP_UV_COR
 - RLE_MED_GC
 
In addition, we also simulate data, and use the following evaluation metrics:

 1. Mock analysis using simulation:
  - False positive rate: fraction of DA peaks at 5% level.
  - p-value distribution difference from uniform distribution, using the Hellinger distance.
  - Is p-value uniformity affected by GC-content: variability in this distance with respect to different GC-content bins.

 2. DA analysis with signal, using simulation:
  - AUROC
  - Impact of GC on DA: how different is the GC-content distribution of DA peaks from the GC-content distribution of truly DA peaks? This again uses the Hellinger distance.
  
# Import and process scone evaluation

Note that the scone scores have already been multiplied by (+/-)1 such that a high score corresponds to a good normalization.

```{r sconeResults}
sconeResultFiles <- list.files("../../objects",  
                               pattern = "*scores.rds$", full.names = TRUE)
sconeResultFilesShort <- list.files("../../objects",  
                               pattern = "*scores.rds$")
datasets <- unlist(lapply(strsplit(sconeResultFilesShort, split="_"),"[[",1))
datasets[datasets %in% "geschwind2018"] <- "torre-ubieta2018"
scores <- sapply(1:length(sconeResultFiles), function(ii) readRDS(sconeResultFiles[ii]))
names(scores) <- datasets
# remove other methods
rmMethods <- c("gcqn_mean", "gcqn_median_10", "gcqn_median_20",  "gcqn_median_100",
               "gcqn_median_permuted")
scores <- lapply(scores, function(scoreMat){
  scoreNames <- rownames(scoreMat)
  rmID <- unlist(sapply(rmMethods, function(method){
    grep(x=scoreNames, pattern=method)
  }))
  return(scoreMat[-rmID,])
})
# remove unused metrics
rmMetrics <- c("RLE_MED", "RLE_IQR", "rleGC_iqr")
scores <- lapply(scores, function(scoreMat){
  scoreColNames <- colnames(scoreMat)
  rmID <- unlist(sapply(rmMetrics, function(metric){
    grep(x=scoreColNames, pattern=metric)
  }))
  return(scoreMat[,-rmID])
})
# focus on no_uv
keepUV <- "no_uv"
scores <- lapply(scores, function(scoreMat){
  scoreNames <- rownames(scoreMat)
  keepID <- grep(x=scoreNames, pattern=keepUV)
  return(scoreMat[keepID,])
})
# rank scores: high rank is good score according to that metric.
rankedScores <- lapply(scores, function(scoreMat){
  apply(scoreMat,2,rank)
})

```

# Import and process simulation evaluation

```{r}
simulationResultFiles <- list.files("../../objects",
                          pattern = "simulationResults", full.names = TRUE)
simulationResultFilesShort <- list.files("../../objects",  
                               pattern = "simulationResults")
datasets <- unlist(lapply(strsplit(simulationResultFilesShort, split="_"),"[[",1))
datasets[datasets %in% "geschwind2018"] <- "torre-ubieta2018"
simulationScores <- sapply(1:length(simulationResultFiles), function(ii){
  readRDS(simulationResultFiles[ii])
}, simplify = FALSE)
names(simulationScores) <- datasets
# remove unused metrics
rmMetricsSim <- "aufdptpr"
simulationScores <- lapply(simulationScores, function(scoreList){
  scoreList[[2]] <- scoreList[[2]][!names(scoreList[[2]]) %in% rmMetricsSim]
  return(scoreList)
})
```

```{r}
# high is good
simulationScores <- lapply(simulationScores, function(simScoreList){
  simScoreList[[1]]$fpr <- -abs(simScoreList[[1]]$fpr - 0.05)
  simScoreList[[1]]$distUnifPvalMock <- -simScoreList[[1]]$distUnifPvalMock
  simScoreList[[1]]$helDistVarGC <- -simScoreList[[1]]$helDistVarGC
  # until fixed:
  simScoreList[[2]]$DAGCDistDiff[is.na(simScoreList[[2]]$DAGCDistDiff)] <- 0
  simScoreList[[2]]$DAGCDistDiff <- -simScoreList[[2]]$DAGCDistDiff
  return(simScoreList)
})
simulationScoreMats <- lapply(simulationScores, function(simScoreList){
  oneList <- c(simScoreList[[1]], simScoreList[[2]])
  # ensure same order
  nameOrder <- names(oneList[[1]])
  oneList <- lapply(oneList, function(x) x[nameOrder])
  scoreMat <- t(do.call(rbind, oneList))
  return(scoreMat)
})
simulationScoreRanks <- lapply(simulationScoreMats, function(scoreMat){
  apply(scoreMat,2,rank)
})
```


# Merge scores

```{r}
rankedScores <- lapply(rankedScores, function(scoreMat){
  rownames(scoreMat) <- unlist(lapply(strsplit(rownames(scoreMat), split=","), "[[", 2))
  rownames(scoreMat)[rownames(scoreMat) == "gcqn_median_50"] <- "gcqn"
  rownames(scoreMat)[rownames(scoreMat) == "fq"] <- "FQ"
  if("cqn_length" %in% rownames(scoreMat)){
    scoreMat <- scoreMat[!rownames(scoreMat) == "cqn",]
    rownames(scoreMat)[rownames(scoreMat) == "cqn_length"] <- "cqn"
  }
  rownames(scoreMat)[rownames(scoreMat) == "tmm"] <- "TMM"
  rownames(scoreMat)[rownames(scoreMat) == "deseq"] <- "DESeq2"
  return(scoreMat)
})

# available datasets
dats <- names(simulationScoreRanks)
allScoreRanks <- list()
for(dd in 1:length(dats)){
  sconeScores <- rankedScores[[dats[dd]]]
  simScores <- simulationScoreRanks[[dats[dd]]]
  sconeScores <- sconeScores[rownames(simScores),]
  allScores <- cbind(sconeScores, simScores)
  allScoreRanks[[dd]] <- allScores
}

for(dd in 1:length(allScoreRanks)){
  curScores <- allScoreRanks[[dd]]
  rn <- rownames(curScores)
  rn[rn == "gcqn"] <- "GC-FQ"
  rn[rn == "edaseq"] <- "FQ-FQ"
  rn[rn == "gcqn_smooth"] <- "GC-FQ_smooth"
  rn[rn == "uq"] <- "UQ"
  rn[rn == "none"] <- "None"
  rn[rn == "sum"] <- "Sum"
  rownames(curScores) <- rn
  allScoreRanks[[dd]] <- curScores
}
```

# Performance visualization

## For each dataset

```{r}
library(ggplot2)
gcNormMethods <- c("cqn", "FQ-FQ", "GC-FQ", "GC-FQ_smooth")
pDat <- list()
for(dd in 1:length(dats)){
  curScores <- allScoreRanks[[dd]]
  curScores <- curScores[order(rowMeans(curScores)),]
  df <- data.frame(rank = c(curScores),
                   method = rep(rownames(curScores), ncol(curScores)),
                   metric = rep(colnames(curScores), each=nrow(curScores)))
  df$method <- factor(df$method, levels = rownames(curScores))
  # introduce class on what metrics assess
  df$class <- NA
  df$class[df$metric %in% c("BIO_SIL", "BATCH_SIL", "PAM_SIL", "EXP_QC_COR", "EXP_UV_COR")] <- "norm"
  df$class[df$metric %in% c("rleGC_med", "helDistVarGC", "DAGCDistDiff")] <- "GC"
  df$class[df$metric %in% c("fpr", "distUnifPvalMock", "auroc", "aufdptpr")] <- "DAA"
  # shorter name for metrics
  df$metric <- as.character(df$metric)
  df$metric[df$metric == "BATCH_SIL"] <- "Batch Sil"
  df$metric[df$metric == "BIO_SIL"] <- "Bio Sil"
  df$metric[df$metric == "DAGCDistDiff"] <- "GC-dist DA"
  df$metric[df$metric == "distUnifPvalMock"] <- "p-val unif"
  df$metric[df$metric == "EXP_QC_COR"] <- "QC cor"
  df$metric[df$metric == "EXP_UV_COR"] <- "UV cor"
  df$metric[df$metric == "helDistVarGC"] <- "p-val GC"
  df$metric[df$metric == "PAM_SIL"] <- "PAM Sil"
  df$metric[df$metric == "rleGC_med"] <- "RLE GC"
  df$dataset <- dats[dd]
  if(dd == 1){
    dfAll <- df
  } else {
    dfAll <- rbind(dfAll, df)
  }
  
  
  dfNorm <- df[df$class == "norm",]
  pNorm <- ggplot(dfNorm, aes(x=metric, y=method, fill=rank)) +
    geom_tile() +
    scale_fill_gradient2() +
  theme(legend.position = "none",
          panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, size=9),
        axis.text.y = element_text(size=12, colour=ifelse(levels(df$method) %in% gcNormMethods,
                                                 "dodgerblue", "black")),
        axis.ticks.y = element_blank()) +
    scale_x_discrete(position = "top") +
    ylab(NULL) +
    xlab("Normalization \n performance") #+
    # annotate("text", x = -0.2, y = 12, label = dats[dd], size=6) +
    # coord_cartesian(ylim = c(0,10), xlim=c(1,length(unique(dfNorm$metric))), clip = "off")
    
  dfDA <- df[df$class == "DAA",]
  pDA <- ggplot(dfDA, aes(x=metric, y=method, fill=rank)) +
    geom_tile() +
    scale_fill_gradient2(low = "skyblue", high="steelblue") +
    theme(legend.position = "none",
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, size=9)) +
    scale_x_discrete(position = "top") +
    xlab("Differential analysis \n performance") +
    ylab(NULL)
  
  dfGC <- df[df$class == "GC",]
  pGC <- ggplot(dfGC, aes(x=metric, y=method, fill=rank)) +
    geom_tile() +
    scale_fill_gradient2(low = "darkseagreen2", high="darkseagreen4") +
    theme(legend.position = "none",
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, size=9)) +
    scale_x_discrete(position = "top") +
    xlab("GC-content bias \n removal") +
    ylab(NULL)
  
  pDat[[dd]] <- cowplot::plot_grid(pNorm, pDA, pGC,
                     nrow = 1,
                     rel_widths = c(1.5, 1, .85))
}

cowplot::plot_grid(plotlist = pDat,
                   nrow = 4, ncol = 2,
                   labels=dats,
                   label_size = 11,
                   label_x = -0.025)
ggsave("~/Dropbox/research/atacseq/bulk/plots/totalBenchmarkPerDataset.pdf",
       height = 14, width = 12)

```

## Across all datasets

```{r}
library(tidyverse)
# get order across all
avNormRanks <- dfAll %>%
  group_by(method) %>%
  summarize(avRank = mean(rank, na.rm=TRUE))
orderedMethods <- as.character(avNormRanks$method[order(avNormRanks$avRank, decreasing=FALSE)])
gcNormMethods <- c("cqn", "FQ-FQ", "GC-FQ", "GC-FQ_smooth")

dfAllNorm <- dfAll[dfAll$class == "norm",]
dfAllNormSummary <- dfAllNorm %>% 
  group_by(dataset, method) %>%
  summarize(meanRank = mean(rank))
dfAllNormSummary$method <- factor(dfAllNormSummary$method, levels=orderedMethods)
dfAllNormSummary$dataset <- factor(dfAllNormSummary$dataset,
                                   levels = c("bryois2018", "calderon2019", "fullard2019",
                                              "murphy2019","torre-ubieta2018",
                                              "philip2017", "rizzardi2019",
                                               "liu2019"))

pNormAll <- ggplot(dfAllNormSummary, aes(x=dataset, y=method, fill=meanRank)) +
    geom_tile() +
    scale_fill_gradient2() +
  theme(legend.position = "none",
          panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, size=9, vjust=.8, hjust=.35),
        axis.text.y = element_text(size = 14, 
                                   colour=ifelse(orderedMethods %in% gcNormMethods,
                                                 "dodgerblue", "black"))) +
    scale_x_discrete(position = "top") +
    ylab(NULL) +
    xlab("Normalization \n performance")

## Differential accessibility analysis
dfAllNorm <- dfAll[dfAll$class == "DAA",]
dfAllNormSummary <- dfAllNorm %>% 
  group_by(dataset, method) %>%
  summarize(meanRank = mean(rank))
dfAllNormSummary$method <- factor(dfAllNormSummary$method, levels=orderedMethods)
dfAllNormSummary$dataset <- factor(dfAllNormSummary$dataset,
                                   levels = c("bryois2018", "calderon2019", "fullard2019",
                                              "murphy2019","torre-ubieta2018",
                                              "philip2017", "rizzardi2019",
                                               "liu2019"))

pDAAll <- ggplot(dfAllNormSummary, aes(x=dataset, y=method, fill=meanRank)) +
    geom_tile() +
    scale_fill_gradient2(low = "skyblue", high="steelblue") +
    theme(legend.position = "none",
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, size=9, vjust=.8, hjust=.35)) +
    scale_x_discrete(position = "top") +
    xlab("Differential analysis \n performance") +
    ylab(NULL)


## GC-content bias removal
dfAllNorm <- dfAll[dfAll$class == "GC",]
dfAllNormSummary <- dfAllNorm %>% 
  group_by(dataset, method) %>%
  summarize(meanRank = mean(rank))
# get order
avNormRanks <- dfAllNormSummary %>%
  group_by(method) %>%
  summarize(avRank = mean(meanRank))
orderedMethods <- as.character(avNormRanks$method[order(avNormRanks$avRank, decreasing=FALSE)])
dfAllNormSummary$method <- factor(dfAllNormSummary$method, levels=orderedMethods)
dfAllNormSummary$dataset <- factor(dfAllNormSummary$dataset,
                                   levels = c("bryois2018", "calderon2019", "fullard2019",
                                              "murphy2019","torre-ubieta2018",
                                              "philip2017", "rizzardi2019",
                                               "liu2019"))

pGCAll <- ggplot(dfAllNormSummary, aes(x=dataset, y=method, fill=meanRank)) +
    geom_tile() +
    scale_fill_gradient2(low = "darkseagreen2", high="darkseagreen4") +
    theme(legend.position = "none",
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, size=9, vjust=.8, hjust=.35)) +
    scale_x_discrete(position = "top") +
    xlab("GC-content bias \n removal") +
    ylab(NULL)

cowplot::plot_grid(pNormAll, pDAAll, pGCAll,
                     nrow = 1,
                     rel_widths = c(1.5, 1, 1))

ggsave("~/Dropbox/research/atacseq/bulk/plots/totalBenchmarkSummary.pdf",
       height = 6, width = 12)
```

## Across all datasets, no GC metrics

```{r}
library(tidyverse)
# get order across all
dfAll_noGC <- dfAll[!dfAll$class == "GC",]
avNormRanks <- dfAll_noGC %>%
  group_by(method) %>%
  summarize(avRank = mean(rank, na.rm=TRUE))
orderedMethods <- as.character(avNormRanks$method[order(avNormRanks$avRank, decreasing=FALSE)])
gcNormMethods <- c("cqn", "FQ-FQ", "GC-FQ", "GC-FQ_smooth")

dfAllNorm <- dfAll_noGC[dfAll_noGC$class == "norm",]
dfAllNormSummary <- dfAllNorm %>% 
  group_by(dataset, method) %>%
  summarize(meanRank = mean(rank))
dfAllNormSummary$method <- factor(dfAllNormSummary$method, levels=orderedMethods)
dfAllNormSummary$dataset <- factor(dfAllNormSummary$dataset,
                                   levels = c("bryois2018", "calderon2019", "fullard2019",
                                              "murphy2019","torre-ubieta2018",
                                              "philip2017", "rizzardi2019",
                                               "liu2019"))

pNormAll <- ggplot(dfAllNormSummary, aes(x=dataset, y=method, fill=meanRank)) +
    geom_tile() +
    scale_fill_gradient2() +
  theme(legend.position = "none",
          panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, size=9, vjust=.8, hjust=.35),
        axis.text.y = element_text(size = 14, 
                                   colour=ifelse(orderedMethods %in% gcNormMethods,
                                                 "dodgerblue", "black"))) +
    scale_x_discrete(position = "top") +
    ylab(NULL) +
    xlab("Normalization \n performance")

## Differential accessibility analysis
dfAllNorm <- dfAll_noGC[dfAll_noGC$class == "DAA",]
dfAllNormSummary <- dfAllNorm %>% 
  group_by(dataset, method) %>%
  summarize(meanRank = mean(rank))
dfAllNormSummary$method <- factor(dfAllNormSummary$method, levels=orderedMethods)
dfAllNormSummary$dataset <- factor(dfAllNormSummary$dataset,
                                   levels = c("bryois2018", "calderon2019", "fullard2019",
                                              "murphy2019","torre-ubieta2018",
                                              "philip2017", "rizzardi2019",
                                               "liu2019"))

pDAAll <- ggplot(dfAllNormSummary, aes(x=dataset, y=method, fill=meanRank)) +
    geom_tile() +
    scale_fill_gradient2(low = "skyblue", high="steelblue") +
    theme(legend.position = "none",
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, size=9, vjust=.8, hjust=.35)) +
    scale_x_discrete(position = "top") +
    xlab("Differential analysis \n performance") +
    ylab(NULL)



cowplot::plot_grid(pNormAll, pDAAll,
                     nrow = 1,
                     rel_widths = c(1.5, 1))

ggsave("~/Dropbox/research/atacseq/bulk/plots/totalBenchmarkSummary_noGC.pdf",
       height = 6, width = 10)
```

## For each benchmark component separately

### Normalization

```{r}
# get order across all
dfAllOnlyNorm <- dfAll[dfAll$class == "norm",]
avNormRanks <- dfAllOnlyNorm %>%
  group_by(method) %>%
  summarize(avRank = mean(rank, na.rm=TRUE))
orderedMethods <- as.character(avNormRanks$method[order(avNormRanks$avRank, decreasing=FALSE)])

dfAllNormSummary <- dfAllOnlyNorm %>% 
  group_by(dataset, method) %>%
  summarize(meanRank = mean(rank))
dfAllNormSummary$method <- factor(dfAllNormSummary$method, levels=orderedMethods)
dfAllNormSummary$dataset <- factor(dfAllNormSummary$dataset,
                                   levels = c("bryois2018", "calderon2019", "fullard2019",
                                              "murphy2019","torre-ubieta2018",
                                              "philip2017", "rizzardi2019",
                                               "liu2019"))

pNormOnly <- ggplot(dfAllNormSummary, aes(x=dataset, y=method, fill=meanRank)) +
    geom_tile() +
    scale_fill_gradient2() +
  theme(legend.position = "none",
          panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, size=9, vjust=.8, hjust=.35),
        axis.text.y = element_text(size = 14, 
                                   colour=ifelse(orderedMethods %in% gcNormMethods,
                                                 "dodgerblue", "black"))) +
    scale_x_discrete(position = "top") +
    ylab(NULL) +
    xlab("Normalization \n performance")
pNormOnly
```

### Differential accessibility analysis

```{r}
# get order across all
dfAllOnlyDA <- dfAll[dfAll$class == "DAA",]
avNormRanks <- dfAllOnlyDA %>%
  group_by(method) %>%
  summarize(avRank = mean(rank, na.rm=TRUE))
orderedMethods <- as.character(avNormRanks$method[order(avNormRanks$avRank, decreasing=FALSE)])

dfAllNormSummary <- dfAllOnlyDA %>% 
  group_by(dataset, method) %>%
  summarize(meanRank = mean(rank))
dfAllNormSummary$method <- factor(dfAllNormSummary$method, levels=orderedMethods)
dfAllNormSummary$dataset <- factor(dfAllNormSummary$dataset,
                                   levels = c("bryois2018", "calderon2019", "fullard2019",
                                              "murphy2019","torre-ubieta2018",
                                              "philip2017", "rizzardi2019",
                                               "liu2019"))

pDAOnly <- ggplot(dfAllNormSummary, aes(x=dataset, y=method, fill=meanRank)) +
    geom_tile() +
    scale_fill_gradient2(low = "skyblue", high="steelblue") +
  theme(legend.position = "none",
          panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, size=9, vjust=.8, hjust=.35),
        axis.text.y = element_text(size = 14, 
                                   colour=ifelse(orderedMethods %in% gcNormMethods,
                                                 "dodgerblue", "black"))) +
    scale_x_discrete(position = "top") +
    ylab(NULL) +
    xlab("Differential analysis \n performance")
pDAOnly
```

### GC-content bias removal

```{r}
# get order across all
dfAllOnlyGC <- dfAll[dfAll$class == "GC",]
avNormRanks <- dfAllOnlyGC %>%
  group_by(method) %>%
  summarize(avRank = mean(rank, na.rm=TRUE))
orderedMethods <- as.character(avNormRanks$method[order(avNormRanks$avRank, decreasing=FALSE)])

dfAllNormSummary <- dfAllOnlyGC %>% 
  group_by(dataset, method) %>%
  summarize(meanRank = mean(rank))
dfAllNormSummary$method <- factor(dfAllNormSummary$method, levels=orderedMethods)
dfAllNormSummary$dataset <- factor(dfAllNormSummary$dataset,
                                   levels = c("bryois2018", "calderon2019", "fullard2019",
                                              "murphy2019","torre-ubieta2018",
                                              "philip2017", "rizzardi2019",
                                               "liu2019"))

pGCOnly <- ggplot(dfAllNormSummary, aes(x=dataset, y=method, fill=meanRank)) +
    geom_tile() +
    scale_fill_gradient2(low = "darkseagreen2", high="darkseagreen4") +
  theme(legend.position = "none",
          panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 25, size=9, vjust=.8, hjust=.35),
        axis.text.y = element_text(size = 14, 
                                   colour=ifelse(orderedMethods %in% gcNormMethods,
                                                 "dodgerblue", "black"))) +
    scale_x_discrete(position = "top") +
    ylab(NULL) +
    xlab("GC-content bias \n removal")
pGCOnly
```

```{r}
cowplot::plot_grid(pNormOnly,
                   pDAOnly,
                   pGCOnly,
                   labels=letters[1:3],
                   nrow = 1)
ggsave("~/Dropbox/research/atacseq/bulk/plots/totalBenchmark_perComponent.pdf",
       width = 12, height = 6)
```

